---
title: "Consumer Spending on Essentials Increased Through 2024"
author: "Mike Konczal"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  message: false
  warning: false
---

```{r load_data}
library(tidyverse)

cex <- read_csv("data/cex_data.csv")
```

## Introduction

The Bureau of Labor Statistics Consumer Expenditure Survey (BLS CEX) is a nationally representative survey of consumer units that reports annual spending by detailed category. It is the primary federal source for household expenditure patterns and is widely used to understand cost burdens, to inform CPI weights, and to benchmark how budgets shift over time. The CEX makes it possible to compare budget shares across time and across demographic groups such as income quintiles and the age of the reference person.

The latest summary statistics for 2024 came out in December 2025, delayed because of the government shutdown. This data helps explain the "vibescession" and other reasons why people's experiences of the economy have been poor even when aggregate income growth keeps pace with inflation. We can see that a rising share of budgets devoted to necessities can make households feel worse off.

## Summary

This report checks whether core essentials and essentials have become a larger share of total expenditures. It compares:

-   2019 vs 2024
-   the 2016-2019 average vs the 2023-2024 average

Results are shown for overall consumer units, the available age category, and income quintiles.

## Method

Core essentials are defined as Food at home and Housing. Essentials add Healthcare and Transportation to the core set. Budget shares are computed by summing category expenditures and dividing by total expenditures ("Average annual expenditures") for each group across years.

```{r essentials_setup}
core_essentials <- c("Food at home", "Housing")
essentials <- c(core_essentials, "Healthcare", "Transportation")

totals <- cex %>%
  filter(item_text == "Average annual expenditures") %>%
  select(year, demographics_text, characteristics_text, total = value)

build_share <- function(items, label) {
  cex %>%
    filter(item_text %in% items) %>%
    group_by(year, demographics_text, characteristics_text) %>%
    summarise(essentials_value = sum(value, na.rm = TRUE), .groups = "drop") %>%
    left_join(totals, by = c("year", "demographics_text", "characteristics_text")) %>%
    mutate(
      essentials_share = essentials_value / total,
      essentials_type = label
    )
}

essentials_share <- bind_rows(
  build_share(core_essentials, "Core essentials"),
  build_share(essentials, "Essentials")
)

group_definitions <- tibble(
  group = c(
    "Overall",
    "Age 25-34",
    "Income: Lowest 20%",
    "Income: Second 20%",
    "Income: Third 20%",
    "Income: Fourth 20%",
    "Income: Highest 20%"
  ),
  demographics_text = c(
    "Quintiles of income before taxes",
    "Age of reference person",
    rep("Quintiles of income before taxes", 5)
  ),
  characteristics_text = c(
    "All Consumer Units",
    "Reference person from age 25 to 34",
    "Lowest 20 percent income quintile",
    "Second 20 percent income quintile",
    "Third 20 percent income quintile",
    "Fourth 20 percent income quintile",
    "Highest 20 percent income quintile"
  )
)

summarize_group <- function(demo_label, char_filter) {
  essentials_share %>%
    filter(
      demographics_text == demo_label,
      characteristics_text %in% char_filter
    ) %>%
    group_by(essentials_type, characteristics_text) %>%
    summarise(
      share_2019 = essentials_share[year == 2019][1],
      share_2024 = essentials_share[year == 2024][1],
      change_2019_2024 = share_2024 - share_2019,
      avg_2016_2019 = mean(essentials_share[year %in% 2016:2019], na.rm = TRUE),
      avg_2023_2024 = mean(essentials_share[year %in% 2023:2024], na.rm = TRUE),
      change_2016_2019_to_2023_2024 = avg_2023_2024 - avg_2016_2019,
      .groups = "drop"
    )
}

format_table <- function(df) {
  df %>%
    mutate(across(where(is.double), ~ scales::percent(.x, accuracy = 0.1))) %>%
    rename(
      type = essentials_type,
      group = characteristics_text,
      `2019` = share_2019,
      `2024` = share_2024,
      `chg 19-24` = change_2019_2024,
      `avg 16-19` = avg_2016_2019,
      `avg 23-24` = avg_2023_2024,
      `chg avg` = change_2016_2019_to_2023_2024
    )
}
```

## Findings

### Overall (All Consumer Units)

```{r overall_results}
overall <- summarize_group(
  "Quintiles of income before taxes",
  "All Consumer Units"
)

overall %>%
  format_table() %>%
  select(type, `2019`, `2024`, `chg 19-24`, `avg 16-19`, `avg 23-24`, `chg avg`) %>%
  knitr::kable()
```

Essentials are slightly higher in 2024 than 2019 for overall consumer units, and the 2023-2024 average is modestly above 2016-2019.

### Age (Reference person from age 25 to 34)

```{r age_results}
age <- summarize_group(
  "Age of reference person",
  "Reference person from age 25 to 34"
)

age %>%
  format_table() %>%
  select(type, `2019`, `2024`, `chg 19-24`, `avg 16-19`, `avg 23-24`, `chg avg`) %>%
  knitr::kable()
```

It seems like this won't hold for the young and anxious, here ages 25-34. So we should exclude this from the analysis.

### Income Quintiles

```{r income_results}
income <- summarize_group(
  "Quintiles of income before taxes",
  c(
    "Lowest 20 percent income quintile",
    "Second 20 percent income quintile",
    "Third 20 percent income quintile",
    "Fourth 20 percent income quintile",
    "Highest 20 percent income quintile"
  )
)

income_order <- c(
  "Lowest 20 percent income quintile",
  "Second 20 percent income quintile",
  "Third 20 percent income quintile",
  "Fourth 20 percent income quintile",
  "Highest 20 percent income quintile"
)

income %>%
  mutate(characteristics_text = factor(characteristics_text, levels = income_order)) %>%
  arrange(characteristics_text, essentials_type) %>%
  format_table() %>%
  select(group, type, `2019`, `2024`, `chg 19-24`, `avg 16-19`, `avg 23-24`, `chg avg`) %>%
  knitr::kable()
```

The lowest two income quintiles show the largest increases since 2019, while the highest quintile is flat to slightly down. The overall pattern suggests an increase in cost burden for essentials concentrated in the lower-middle of the income distribution. Here it is with details, for the years 2019 and 2024, as well as the average across 2016-2019 and 2023-2024.

In addition are bar charts and graphics showing the increase.

## Core Essentials Visuals

```{r core_bar, fig.height=7.5}
core_share <- essentials_share %>%
  filter(essentials_type == "Core essentials") %>%
  inner_join(group_definitions, by = c("demographics_text", "characteristics_text"))

core_bar <- core_share %>%
  filter(year %in% c(2016:2019, 2023:2024)) %>%
  mutate(period = if_else(year <= 2019, "2016-2019 avg", "2023-2024 avg")) %>%
  mutate(period = factor(period, levels = c("2016-2019 avg", "2023-2024 avg"))) %>%
  group_by(group, period) %>%
  summarise(avg_share = mean(essentials_share, na.rm = TRUE), .groups = "drop")

ggplot(
  core_bar,
  aes(x = factor(group, levels = group_definitions$group), y = avg_share, fill = period)
) +
  geom_col(position = position_dodge2(width = 0.7, reverse = TRUE), width = 0.65) +
  geom_text(
    aes(label = scales::percent(avg_share, accuracy = 0.1)),
    position = position_dodge2(width = 0.7, reverse = TRUE),
    hjust = 1.05,
    color = "white",
    size = 3.2
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("2016-2019 avg" = "#9EC7E0", "2023-2024 avg" = "#1F5A7A")) +
  labs(
    title = "Core essentials share: pre-pandemic vs 2023-2024",
    x = NULL,
    y = "Core essentials share",
    fill = NULL
  ) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
```

```{r core_line, fig.height=7.5}
core_line <- core_share %>%
  filter(year >= 2010) %>%
  mutate(group = factor(group, levels = group_definitions$group))

ggplot(core_line, aes(x = year, y = essentials_share, group = group)) +
  geom_line(color = "#1F5A7A", linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~ group, ncol = 2, scales = "free_y") +
  labs(
    title = "Core essentials share by group, 2010-2024",
    x = NULL,
    y = "Core essentials share"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Conclusion

The 2020-2024 period marks a clear upward shift in core essentials, especially for lower-income households, consistent with a "vibescession" narrative where everyday bills rise faster than perceptions of income gains. Policy responses that reduce housing costs, stabilize food prices, and expand access to affordable healthcare and transportation would directly address the areas where cost burdens are rising most, while also speaking to the lived experience of people who find themselves struggling in this economy.
