---
title: "Phillips Curves With Alternative Augmentations"
format:
  html:
    toc: true
    code-fold: show
date: "`r Sys.Date()`"
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(lubridate)
library(janitor)
library(zoo)
library(modelsummary)
library(broom)
library(gt)
library(texreg)
library(huxtable)
library(kableExtra)

options(dplyr.summarise.inform = FALSE)
```

**Data sources and download approach**
- Core PCE inflation, CPI inflation, unemployment rate, NAIRU, unemployed level, and job openings come from FRED via the public `fredgraph.csv` endpoint.
- Expected inflation (10-year CPI) comes from the Survey of Professional Forecasters (SPF) mean responses Excel file.
- GSCPI comes from the NY Fed monthly data file (cached locally as CSV when available).
- YouTube views are pulled from the Kworb monthly views table for the official “good 4 u” video.
All downloads and the dataset assembly are handled by `/Users/mtkonczal/Documents/good4u_inflation/scripts/build_phillips_data.py`, which writes a single quarterly CSV at `data/phillips_data.csv` used below.

**What this Quarto does**
- Downloads monthly CPI, unemployment, unemployed, and job openings from FRED; SPF 10-year expected CPI inflation from the Philadelphia Fed; and GSCPI from the NY Fed.
- Aggregates all series to quarterly frequency. Core PCE inflation is computed as quarterly log differences annualized: \(\pi^{core}_t = 400(\ln P^{core}_t - \ln P^{core}_{t-1})\).
- Builds Phillips Curve regressions with expected inflation and two lags of realized inflation, then augments the curve with one variable at a time and with all variables together.
- Uses the unemployment gap \(u_t - u^*_t\) (unemployment minus NAIRU).

**Equations**

Baseline Phillips Curve (quarterly):

\[
\pi^{core}_t = \beta_0 + \beta_1 \mathbb{E}_t[\pi^{10y}] + \beta_2 \pi^{core}_{t-1} + \beta_3 \pi^{core}_{t-2} + \beta_4 (u_t - u^*_t) + \varepsilon_t
\]

Augmented versions (one at a time):

\[
\pi^{core}_t = \beta_0 + \beta_1 \mathbb{E}_t[\pi^{10y}] + \beta_2 \pi^{core}_{t-1} + \beta_3 \pi^{core}_{t-2} + \beta_4 (u_t - u^*_t) + \gamma Z_t + \varepsilon_t
\]

where \(Z_t\) is one of: GSCPI, \(v/u\), \(\log(1 + \text{Views})\) for Good 4 U, or \(1\{2021 \le t \le 2022\}\).

All-augments model:

\[
\pi^{core}_t = \beta_0 + \beta_1 \mathbb{E}_t[\pi^{10y}] + \beta_2 \pi^{core}_{t-1} + \beta_3 \pi^{core}_{t-2} + \beta_4 (u_t - u^*_t) + \gamma_1 \text{GSCPI}_t + \gamma_2 (v/u)_t + \gamma_3 \log(1+\text{Views}_t) + \gamma_4 D_{2021\text{–}2022} + \varepsilon_t
\]

```{r}
#| label: helpers
load_phillips_data <- function(path) {
  if (!file.exists(path)) {
    stop("Missing data/phillips_data.csv. Run scripts/build_phillips_data.py first.")
  }

  readr::read_csv(path, show_col_types = FALSE) |>
    mutate(qtr = as.yearqtr(as.Date(qtr_date)))
}
```

**Load data**

```{r}
#| label: load-data
phillips_df <- load_phillips_data("data/phillips_data.csv")
```

**Build quarterly dataset**

```{r}
#| label: build-quarterly
phillips_df <- phillips_df |>
  filter(!is.na(core_pce_pi), !is.na(unrate_gap), !is.na(exp_pi10)) |>
  filter(as.integer(format(as.Date(qtr), "%Y")) >= 2000)

sample_range <- range(phillips_df$qtr, na.rm = TRUE)
```

```{r}
#| label: sample-note
cat("Sample range:", format(sample_range[1]), "to", format(sample_range[2]), "\n")
```

**Model specifications**

```{r}
#| label: models

base_formula <- core_pce_pi ~ exp_pi10 + core_pce_pi_l1 + core_pce_pi_l2 + unrate_gap

fit_base <- lm(base_formula, data = phillips_df)
fit_gscpi <- lm(update(base_formula, . ~ . + gscpi), data = phillips_df)
fit_vu <- lm(update(base_formula, . ~ . + v_u), data = phillips_df)

have_youtube <- any(!is.na(phillips_df$good4u_views_log))
fit_youtube <- if (have_youtube) {
  lm(update(base_formula, . ~ . + good4u_views_log), data = phillips_df)
} else {
  NULL
}

fit_dummy <- lm(update(base_formula, . ~ . + y2021_2022), data = phillips_df)

fit_all <- if (have_youtube) {
  lm(update(base_formula, . ~ . + gscpi + v_u + good4u_views_log + y2021_2022), data = phillips_df)
} else {
  NULL
}

models <- list(
  "Baseline" = fit_base,
  "GSCPI" = fit_gscpi,
  "v/u" = fit_vu,
  "Good 4 U views (log1p)" = fit_youtube,
  "2021-2022 dummy" = fit_dummy,
  "All four" = fit_all
)

models <- models[!vapply(models, is.null, logical(1))]
```

**Regression table**

```{r}
#| label: regression-table

coef_tbl <- imap_dfr(models, function(m, name) {
  broom::tidy(m) |>
    mutate(model = name)
}) |>
  filter(term != "(Intercept)") |>
  mutate(
    stars = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ ""
    ),
    term = recode(
      term,
      exp_pi10 = "Expected inflation (10y)",
      core_pce_pi_l1 = "Core PCE inflation lag 1",
      core_pce_pi_l2 = "Core PCE inflation lag 2",
      unrate_gap = "Unemployment gap (u - u*)",
      gscpi = "GSCPI",
      v_u = "v/u",
      good4u_views_log = "Good 4 U views (log1p)",
      y2021_2022 = "2021-2022 dummy"
    ),
    estimate_fmt = sprintf("%.3f%s", estimate, stars),
    se_fmt = sprintf("(%.3f)", std.error)
  )

term_levels <- coef_tbl |>
  distinct(term) |>
  pull(term)

est_wide <- coef_tbl |>
  select(term, model, estimate_fmt) |>
  pivot_wider(names_from = model, values_from = estimate_fmt) |>
  mutate(term_order = match(term, term_levels), row_type = 1)

se_wide <- coef_tbl |>
  select(term, model, se_fmt) |>
  pivot_wider(names_from = model, values_from = se_fmt) |>
  mutate(term_order = match(term, term_levels), row_type = 2, term = "")

reg_table <- bind_rows(est_wide, se_wide) |>
  arrange(term_order, row_type) |>
  select(-term_order, -row_type) |>
  mutate(across(everything(), ~ ifelse(is.na(.), "", .)))

knitr::kable(reg_table, caption = "Phillips Curve regressions (standard errors in parentheses)")
```

**Regression table (stargazer)**

```{r}
#| label: regression-table-stargazer
#| results: 'asis'

cov_labels <- c(
  "Expected inflation (10y)",
  "Core PCE inflation lag 1",
  "Core PCE inflation lag 2",
  "Unemployment gap (u - u*)",
  "GSCPI",
  "v/u",
  "Good 4 U views (log1p)",
  "2021-2022 dummy"
)

model_list <- unname(models)

cat('<div class="stargazer-wide">')
cat('<style>.stargazer-wide table{width:100%;}</style>')
stargazer::stargazer(
  model_list,
  type = "html",
  title = "Phillips Curve regressions (stargazer format)",
  dep.var.labels = "Core PCE inflation (annualized)",
  covariate.labels = cov_labels,
  omit.stat = c("f", "ser"),
  digits = 3,
  column.sep.width = "12pt",
  font.size = "small"
)
cat('</div>')
```

**Regression table (gt via modelsummary)**

```{r}
#| label: regression-table-gt

modelsummary::modelsummary(
  models,
  coef_omit = "Intercept",
  output = "gt"
)
```

**Regression table (texreg)**

```{r}
#| label: regression-table-texreg
#| results: 'asis'

coef_map <- list(
  exp_pi10 = "Expected inflation (10y)",
  core_pce_pi_l1 = "Core PCE inflation lag 1",
  core_pce_pi_l2 = "Core PCE inflation lag 2",
  unrate_gap = "Unemployment gap (u - u*)",
  gscpi = "GSCPI",
  v_u = "v/u",
  good4u_views_log = "Good 4 U views (log1p)",
  y2021_2022 = "2021-2022 dummy"
)

short_model_names <- dplyr::recode(
  names(models),
  "Baseline" = "Base",
  "GSCPI" = "GSCPI",
  "v/u" = "VU",
  "Good 4 U views (log1p)" = "G4U",
  "2021-2022 dummy" = "D2122",
  "All four" = "All4",
  .default = names(models)
)

texreg::htmlreg(
  model_list,
  custom.coef.map = coef_map,
  custom.model.names = short_model_names,
  caption = "Phillips Curve regressions (texreg format)",
  omit.stat = c("f", "ser"),
  digits = 3
)
```

**Regression table (huxtable)**

```{r}
#| label: regression-table-huxtable

huxtable::huxreg(
  model_list,
  coefs = c(
    "Expected inflation (10y)" = "exp_pi10",
    "Core PCE inflation lag 1" = "core_pce_pi_l1",
    "Core PCE inflation lag 2" = "core_pce_pi_l2",
    "Unemployment gap (u - u*)" = "unrate_gap",
    "GSCPI" = "gscpi",
    "v/u" = "v_u",
    "Good 4 U views (log1p)" = "good4u_views_log",
    "2021-2022 dummy" = "y2021_2022"
  )
)
```

**Regression table (kableExtra)**

```{r}
#| label: regression-table-kableextra
#| results: 'asis'

kable_df <- modelsummary::modelsummary(
  models,
  coef_omit = "Intercept",
  output = "data.frame"
) |>
  mutate(across(everything(), ~ ifelse(is.na(.), "", .)))

knitr::kable(kable_df, format = "html", caption = "Phillips Curve regressions (kableExtra)") |>
  kableExtra::kable_styling(full_width = TRUE, fixed_thead = TRUE) |>
  kableExtra::scroll_box(width = "100%", height = "420px")
```
