---
title: "A 4-Good Almost Ideal Demand System for Affordability"
author: "Mike Konczal"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| echo: false
library(tidyverse)
library(gt)
library(scales)
library(systemfit)
```

## Why This Model

A rising budget share for essentials can happen for different reasons. Prices might rise faster than other prices, which forces households to reallocate spending even if preferences are unchanged. Or households might actively choose to buy more of a category as incomes rise. A demand system is useful because it separates those channels inside one coherent framework.

This report estimates a single Linear Approximate Almost Ideal Demand System (LA/AIDS) using four categories only:

- Food at home
- Food away from home
- Shelter
- Everything else

The goal is to measure how each category reacts to total spending and to its own price, then interpret what that implies for affordability.

## How The LA/AIDS Model Works

### Core equation

For each good \(i\), the model explains its budget share \(w_{it}\):

\[
w_{it} = \alpha_i + \sum_j \gamma_{ij} \ln p_{jt} + \beta_i \ln\left(\frac{X_t}{P_t^*}\right) + \delta_i\,\text{trend}_t + \varepsilon_{it}
\]

Where:

- \(w_{it}\) is spending on good \(i\) as a share of total spending in month \(t\)
- \(p_{jt}\) are price indexes for each good
- \(X_t\) is nominal total expenditure
- \(P_t^*\) is the Stone price index, a weighted average of log prices
- \(\beta_i\) governs the expenditure elasticity
- \(\gamma_{ij}\) governs price effects and substitution patterns

### Why this is useful for affordability

The model gives three outputs for each good:

- Expenditure elasticity: how spending on that good moves when total spending rises
- Marshallian own-price elasticity: total quantity response to its own price change
- Hicksian own-price elasticity: substitution-only quantity response, net of income effects

These are the key objects for saying whether a category behaves like a necessity or luxury, and whether households can substitute away when prices rise.

### Restrictions from consumer theory

The AIDS system imposes standard demand restrictions:

- Adding-up: budget shares sum to 1 each month
- Homogeneity: only relative prices matter
- Symmetry: cross-price effects are symmetric after accounting for income effects

Implementation details in this report:

- Everything else is treated as the residual good implied by adding-up
- Prices are expressed relative to the residual good to impose homogeneity
- Three share equations are estimated by SUR; the fourth is implied by adding-up
- Symmetry is imposed through cross-equation restrictions

## Data And Category Construction

The source is monthly BEA PCE data from January 2010 onward (`data/pce_data.csv`).

Category mapping:

| Model category | PCE series |
|---|---|
| Shelter | Housing and utilities |
| Food at home | Food and beverages purchased for off-premises consumption |
| Food away from home | Food services and accommodations |
| Everything else | Residual of total PCE after the three categories |

Residual price index construction:

- Normalize all observed price indexes to 1 at the sample start
- Use the identity for the aggregate log price index to back out the residual log price
- This keeps the full system internally consistent with observed overall inflation

## Estimation

```{r estimate_4good}
# 1) Load and reshape monthly PCE data
pce <- read_csv("data/pce_data.csv") %>%
  distinct(date, item_name, .keep_all = TRUE) %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= as.Date("2010-01-01")) %>%
  arrange(item_name, date)

items_needed <- c(
  "Housing and utilities",
  "Food and beverages purchased for off-premises consumption",
  "Food services and accommodations",
  "Personal consumption expenditures"
)

wide <- pce %>%
  filter(item_name %in% items_needed) %>%
  select(date, item_name, value, PCEweight, quantity) %>%
  pivot_wider(
    id_cols = date,
    names_from = item_name,
    values_from = c(value, PCEweight, quantity)
  ) %>%
  drop_na() %>%
  arrange(date)

names(wide) <- names(wide) %>%
  str_replace_all("Housing and utilities", "shelter") %>%
  str_replace_all("Food and beverages purchased for off-premises consumption", "food_home") %>%
  str_replace_all("Food services and accommodations", "food_away") %>%
  str_replace_all("Personal consumption expenditures", "pce")

# 2) Build shares and prices
wide <- wide %>%
  mutate(
    w_shelter = PCEweight_shelter,
    w_home = PCEweight_food_home,
    w_away = PCEweight_food_away,
    w_other = 1 - w_shelter - w_home - w_away,
    p_shelter = value_shelter / value_shelter[1],
    p_home = value_food_home / value_food_home[1],
    p_away = value_food_away / value_food_away[1],
    p_overall = value_pce / value_pce[1],
    ln_p_shelter = log(p_shelter),
    ln_p_home = log(p_home),
    ln_p_away = log(p_away),
    ln_p_overall = log(p_overall),
    ln_p_other = (
      ln_p_overall - w_shelter * ln_p_shelter - w_home * ln_p_home - w_away * ln_p_away
    ) / w_other,
    p_other = exp(ln_p_other),
    lnp_sh = ln_p_shelter - ln_p_other,
    lnp_h = ln_p_home - ln_p_other,
    lnp_a = ln_p_away - ln_p_other,
    ln_Pstar = w_shelter * ln_p_shelter + w_home * ln_p_home + w_away * ln_p_away + w_other * ln_p_other,
    X = value_pce * quantity_pce,
    lnXP = log(X) - ln_Pstar,
    trend = as.numeric(difftime(date, min(date), units = "days")) / 30.44
  )

# 3) Estimate 3 equations by SUR with symmetry restrictions
aids4_data <- wide %>%
  select(w_shelter, w_home, w_away, lnp_sh, lnp_h, lnp_a, lnXP, trend) %>%
  drop_na()

eq_shelter <- w_shelter ~ lnp_sh + lnp_h + lnp_a + lnXP + trend
eq_home <- w_home ~ lnp_sh + lnp_h + lnp_a + lnXP + trend
eq_away <- w_away ~ lnp_sh + lnp_h + lnp_a + lnXP + trend

system4 <- list(shelter = eq_shelter, home = eq_home, away = eq_away)

symmetry_restrictions <- c(
  "shelter_lnp_h - home_lnp_sh = 0",
  "shelter_lnp_a - away_lnp_sh = 0",
  "home_lnp_a - away_lnp_h = 0"
)

sur4 <- systemfit(
  system4,
  method = "SUR",
  data = aids4_data,
  restrict.matrix = symmetry_restrictions
)

# 4) Extract coefficients
coefs4 <- summary(sur4)$coefficients
coef4_df <- tibble(
  term = rownames(coefs4),
  estimate = coefs4[, "Estimate"],
  std_error = coefs4[, "Std. Error"],
  t_value = coefs4[, "t value"],
  p_value = coefs4[, "Pr(>|t|)"]
)

# 5) Elasticities at sample means
classify <- function(e) {
  case_when(
    e < 0 ~ "Inferior",
    e < 1 ~ "Necessity",
    e == 1 ~ "Unitary",
    TRUE ~ "Luxury"
  )
}

w_means <- colMeans(aids4_data[, c("w_shelter", "w_home", "w_away")])
w_other_mean <- 1 - sum(w_means)

eq_names <- c("shelter", "home", "away")
lnp_names <- c("lnp_sh", "lnp_h", "lnp_a")

betas <- sapply(eq_names, function(eq) coef(sur4)[paste0(eq, "_lnXP")])
gammas_own <- sapply(seq_along(eq_names), function(i) {
  coef(sur4)[paste0(eq_names[i], "_", lnp_names[i])]
})

e_exp <- 1 + betas / w_means
beta_other <- -sum(betas)
e_exp_other <- 1 + beta_other / w_other_mean

e_m <- -1 + gammas_own / w_means - betas

gamma_mat <- matrix(0, length(eq_names), length(eq_names))
for (i in seq_along(eq_names)) {
  for (j in seq_along(eq_names)) {
    gamma_mat[i, j] <- coef(sur4)[paste0(eq_names[i], "_", lnp_names[j])]
  }
}
gamma_i_other <- -rowSums(gamma_mat)
gamma_other_other <- -sum(gamma_i_other)
e_m_other <- -1 + gamma_other_other / w_other_mean - beta_other

e_h <- e_m + w_means * e_exp
e_h_other <- e_m_other + w_other_mean * e_exp_other

elasticities4 <- tibble(
  good = c("Shelter", "Food at Home", "Food Away from Home", "Everything Else"),
  mean_share = c(w_means, w_other_mean),
  expenditure_elasticity = c(e_exp, e_exp_other),
  marshallian_own_price = c(e_m, e_m_other),
  hicksian_own_price = c(e_h, e_h_other),
  classification = classify(c(e_exp, e_exp_other))
)

write_csv(coef4_df, "data/aids4_coefficients.csv")
write_csv(elasticities4, "data/aids4_elasticities.csv")
```

## Coefficient Estimates

```{r coef_table_4good}
coef4_df %>%
  mutate(
    term = term %>%
      str_replace("shelter_", "Shelter: ") %>%
      str_replace("home_", "Food Home: ") %>%
      str_replace("away_", "Food Away: ") %>%
      str_replace("\\(Intercept\\)", "Intercept") %>%
      str_replace("lnp_sh", "ln(p_shelter/p_other)") %>%
      str_replace("lnp_h", "ln(p_home/p_other)") %>%
      str_replace("lnp_a", "ln(p_away/p_other)") %>%
      str_replace("lnXP", "ln(X/P*)") %>%
      str_replace("trend", "Trend")
  ) %>%
  gt() %>%
  tab_header(title = "4-Good LA/AIDS: SUR Coefficient Estimates") %>%
  fmt_number(columns = c(estimate, std_error, t_value), decimals = 4) %>%
  fmt_number(columns = p_value, decimals = 4) %>%
  cols_label(
    term = "Term",
    estimate = "Estimate",
    std_error = "Std. Error",
    t_value = "t",
    p_value = "p-value"
  )
```

## Elasticity Results

```{r elasticity_table_4good}
elasticities4 %>%
  gt() %>%
  tab_header(title = "4-Good AIDS: Elasticities at Sample Means") %>%
  fmt_number(
    columns = c(mean_share, expenditure_elasticity, marshallian_own_price, hicksian_own_price),
    decimals = 3
  ) %>%
  cols_label(
    good = "Good",
    mean_share = "Mean Share",
    expenditure_elasticity = "Expenditure",
    marshallian_own_price = "Marshallian",
    hicksian_own_price = "Hicksian",
    classification = "Type"
  ) %>%
  tab_spanner(label = "Own-Price Elasticity", columns = c(marshallian_own_price, hicksian_own_price))
```

```{r elasticity_plot_4good}
#| fig-width: 8
#| fig-height: 5

ggplot(elasticities4, aes(x = expenditure_elasticity, y = reorder(good, expenditure_elasticity))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_segment(aes(xend = 0, yend = good), color = "#1F5A7A", linewidth = 0.8) +
  geom_point(size = 4, color = "#1F5A7A") +
  scale_x_continuous(breaks = seq(-1, 3, 0.5)) +
  labs(
    title = "Expenditure Elasticities: 4-Good AIDS Model",
    subtitle = "Food at home, food away from home, shelter, everything else",
    x = "Expenditure elasticity",
    y = NULL,
    caption = "Source: BEA PCE, Author's AIDS estimation; Mike Konczal."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 11)
  )

ggsave("graphics/aids4_elasticities.png", width = 8, height = 5, dpi = 150)
```

## Interpretation

This model estimates one coherent system for the four categories, so the categories compete for the same dollar of household spending each month.

- Shelter (`r round(elasticities4$expenditure_elasticity[elasticities4$good == "Shelter"], 2)`) is a necessity. Its share tends to fall slowly as total spending rises, which is exactly why shelter inflation creates persistent pressure.
- Food at home (`r round(elasticities4$expenditure_elasticity[elasticities4$good == "Food at Home"], 2)`) is estimated as inferior. That is consistent with households shifting toward food away from home as budgets expand.
- Food away from home (`r round(elasticities4$expenditure_elasticity[elasticities4$good == "Food Away from Home"], 2)`) is strongly luxury-like. It scales up quickly in expansions.
- Everything else (`r round(elasticities4$expenditure_elasticity[elasticities4$good == "Everything Else"], 2)`) is above unit elasticity in this specification, so it behaves like a broad discretionary bundle.

Some own-price elasticities are weak or signed counter to simple intuition. In a compact system with correlated macro price indexes, that can happen because relative-price movements are limited. The expenditure elasticities are the most stable object here for classifying necessity versus luxury behavior.

## Bottom Line

Using one 4-good AIDS system with the requested categories and the same SUR methodology, affordability pressure is concentrated in shelter and food at home, while food away from home behaves like the most income-sensitive margin of household demand.
